A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-11-05 latex/main.tex
Second para

Things to say.
\begin{itemize}
\item Explain how it works. Post-hoc. Note methodological distinction 
with sc2ts.
\item Comparison at k/p=4 overall.
\item Discussion of X lineages.
\item Discussion of parameter choice and our lower bound on recomb.
\end{itemize}

@@

[Text from Sally]
The main advantage of sc2ts over previous methods to infer recombination
phylogenetically is that recombination events are explicitly handled as the ARG
is built. Other methods (e.g., RIPPLES) first build a phylogeny, assuming no
recombination, and then look for unusual features (e.g., unusually long
branches). Phylogenetic methods that ignore recombination, however, distort the
tree shape (REFS - Schierup and Hein 2000) and force placement of recombinant
lineages in places that minimize the discrepancy across the entire sequence,
rather than near either of the parent lineages.  As a consequence, many true
recombination events can be missed because their branches are artificially
inferred to be short on reconstructed trees that do not incorporate
recombination. In addition, many of the recombinant sequences can be placed
before their parents in time, because recombinants can appear to be more
similar to ancestral sequences that are known with more uncertainty [OR because
recombinants are placed anywhere on a tree that best fits].

This ``parsimony improvement'' parameter ($p$)
is equivalent to the sc2ts recombination penalty ($k$).

RIPPLES~\cite{Turakhia2022} is a method for identifying
recombination events using a phylogeny inferred by UShER.
This distinguishes RIPPLES from other recombination detection
methods such as 3SEQ, which classify particular \emph{sequences}
as recombinant or not and can only identify 
potential onward transmission post-hoc. 
The combination of UShER and RIPPLES
is therefore the only method that is directly comparable to sc2ts. 
RIPPLES works by first identifying long 
branches in the input tree, by default those carrying 3 or more 
mutations. For each candidate branch,
RIPPLES systematically splits the sequence into 2 or 3 segments, and
finds parsimonious placements for each segment independently.

To compare the performance of sc2ts and UShER+RIPPLES 
in terms of recombinant detection, we ran RIPPLES on the Viridian UShER phylogeny
subset to the 2,475,418 samples shared with the sc2ts ARG (STAR Methods).
Note that we are interested in the direct output of sc2ts and UShER+RIPPLES
here, and do not perform QC filtering of either.
Using the RIPPLES parameter $p=4$ (matching $k=4$ in the sc2ts ARG) resulted in 1174 reported recombination events.
We ran 3SEQ on each of the candidate trios (STAR Methods) and 601 
were classified as recombinant, giving an overall validation rate of 51\%.
Aligning the events reported by both methods using the sets of descendant
samples (STAR Methods), 468 of the events are shared by both,
with the intersection comprising 40\% of the RIPPLES events
and 55\% of the sc2ts events (Figure~\ref{fig:ripples_sc2ts_events}). 
The 706 recombination events output by RIPPLES that do not 
correspond to a recombination event in the sc2ts ARG can largely
be explained by a combination of two factors.
Firstly, 408 of these RIPPLES identified recombination events
have one or more parents that are \emph{younger} than the
recombinant sequence based on metadata. Thus, the parental
sequences were simply not available to match against for 
sc2ts (which enforces strict time ordering) when the 
recombinant sequence was inserted. Secondly, of the 179
remaining events with matching clades and in which the 
sc2ts node has more than 4 mutations, of the total
1,167 mutations, 452 of these occurred at sites that were 
remapped during post-hoc parsimony. The remaining ``missing''
events are likely attributable to the finer details of the 
differences in tree building approaches between UShER and sc2ts.

The 387 events detected by sc2ts and not by UShER+RIPPLES
would require a detailed dissection of the UShER tree
to explain, which is beyond the scope of this work.
However, the set does contain some well characterized events such as XBB.
We also compared the recombination events
output by RIPPLES with the Pango X lineages
(Table~\ref{tab:ripples_x_lineages}).
At p=4, we found that only 9 of the Pango X lineages were present and
with unequivocal recombinant origin for 6 (contrast with 
16 Type I events in Table~\ref{tab:pango_x_lineages}).
We therefore reran at p=3, resulting in 4111 events overall,
and increasing the intersection of events with sc2ts to 595 (70\%).
At p=3, events involving XX of the 44 Pango X lineages
are present in the dataset, with unequivocal origins for 10 lineages, and
7 lineages other lineages associated with two competing events.
[More summary]
[Note somewhere here that XB and XAJ are found to be recombinants 
at p=4, showing that UShER is doing better here for some reason]

In terms of \emph{detecting} recombination events, 
sc2ts and the combination of UShER and RIPPLES have much in common, 
and it is likely that very similar results could be obtained if run
in the same way. That is, if we constructed an UShER tree by 
incrementally adding the same daily batches as used for sc2ts
and examined any newly-added long branches each day
using RIPPLES, we would likely get very similar results to sc2ts
in terms of recombinants detected.
For a given matching operation, RIPPLES and the LS HMM 
(under its current parameterisation) 
are deterministically solving the same well-defined problem
(albeit with radically different implementations) and they 
must therefore produce similar answers. 
Where sc2ts and UShER+RIPPLES diverge, however, is in terms 
of how well such signals of recombination can be preserved
over longer timescales. Continuing our thought experiment
in which we run UShER and RIPPLES in daily batches, suppose
that on a given day a branch with 4 mutations is identified 
as being a recombinant. Unless this branch is specifically 
``protected'' as representing a recombinant it may 
be broken up by subsequent parsimony operations, thus 
erasing the signal of recombination. This would appear
to be the mostly likely explanation for the ``missing'' 
X-lineages like XBB (detected by sc2ts at $k=4$ but 
not by UShER+RIPPLES at $p=4$).

@@

@@

\begin{table}
\caption{
RIPPLES events associated with Pango X lineages.
Each row is a recombination event identified by RIPPLEs (at p = 3)
in which Pango X lineages appear,
along with the corresponding sc2ts event from Table~\ref{tab:pango_x_lineages},
where applicable, grouped by class.
Events also present at p = 4 are indicated.
Clades are matched by finding the most recent common ancestor of
the samples associated with each UShER/RIPPLES event in the sc2ts ARG.
The clade differences indicates how close this match is.
}
\label{tab:ripples_x_lineages}
\centering
\begin{scriptsize}

>>>>

@@

To compare the performance of sc2ts and UShER+RIPPLES 
in terms of recombinant detection, we ran RIPPLES on the Viridian UShER phylogeny
subset to the 2,475,418 samples shared with the sc2ts ARG (STAR Methods).
Note that we are interested in the direct output of sc2ts and UShER+RIPPLES
here, and do not perform QC filtering of either.
Using the RIPPLES parameter $p=4$ (matching $k=4$ used in the sc2ts ARG) 
resulted in 1174 reported recombination events.
We ran 3SEQ on each of the candidate trios (STAR Methods) and 601 
were classified as recombinant, giving an overall validation rate of 51\%.
Aligning the events reported by both methods using the sets of descendant
samples (STAR Methods), 468 of the events are shared by both,
with the intersection comprising 40\% of the RIPPLES events
and 55\% of the sc2ts events (Figure~\ref{fig:ripples_sc2ts_events}). 
The 706 recombination events output by RIPPLES that do not 
correspond to a recombination event in the sc2ts ARG can largely
be explained by a combination the choice of sites 
used for inference, and the strict time ordering imposed
by sc2ts (Document~\ref{sec:detailed_ripples}).
The 387 events detected by sc2ts and not by UShER+RIPPLES
would require a detailed dissection of the UShER tree
to explain, which is beyond the scope of this work.
However, the set does contain some well characterized events 
associated with the Pango X lineages.
Table~\ref{tab:ripples_x_lineages} shows the events detected 
by UShER+RIPPLES associated with samples assigned to Pango X
lineages. At the RIPPLES ``parsimony improvement'' parameter
$p=4$ (equivalent to $k=4$ parameter used to build the sc2ts ARG)
only 9 of the Pango X lineages are present and
with unequivocal recombinant origin for 6 
(contrast with 16 Type I events in Table~\ref{tab:pango_x_lineages}).
We therefore reran RIPPLES with $p=3$, resulting in 4111 events overall,
and increasing the intersection of events with sc2ts to 595 (70\%;
Figure~\ref{fig:ripples_sc2ts_events}).
At this more lenient threshold, 23 of the 44 Pango X lineages
are present in the dataset, with unequivocal origins for 10 lineages, and
7 lineages associated with two competing events 
(Document~\ref{sec:detailed_ripples}).
XB and XAJ are among the lineages detected by UShER+RIPPLES,
strengthening evidence that these are false negatives 
for sc2ts (Table~\ref{tab:pango_x_lineages}).

@@

Firstly, 408 of these RIPPLES identified recombination events
have one or more parents that are \emph{younger} than the
recombinant sequence based on metadata. Thus, the parental
sequences were simply not available to match against for 
sc2ts (which enforces strict time ordering) when the 
recombinant sequence was inserted. Secondly, of the 179
remaining events with matching clades and in which the 
sc2ts node has more than 4 mutations, of the total
1,167 mutations, 452 of these occurred at sites that were 
remapped during post-hoc parsimony. The remaining ``missing''
events are likely attributable to the finer details of the 
differences in tree building approaches between UShER and sc2ts

@@

\begin{table}
\caption{
RIPPLES events associated with Pango X lineages.
Each row is a recombination event identified by RIPPLEs (at p = 3)
in which Pango X lineages appear,
along with the corresponding sc2ts event from Table~\ref{tab:pango_x_lineages},
where applicable, grouped by class.
Events also present at p = 4 are indicated.
Clades are matched by finding the most recent common ancestor of
the samples associated with each UShER+RIPPLES event in the sc2ts ARG.
The clade differences indicates how close this match is.
}
\label{tab:ripples_x_lineages}
\centering
\begin{scriptsize}
